---
title: "Converting a document with doclr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Converting a document with doclr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`doclr` converts documents by talking to a
[docling-serve](https://docling-project.github.io/docling/usage/api_server/rest_api/)
instance over HTTP. The models run on the server; your R session only sends
requests and reads results back. Every chunk below is marked `eval = FALSE`,
because building this vignette should not require a running server.

```{r setup}
library(doclr)
```

## Getting a server

If someone already runs docling-serve for you, all you need is its URL:

```{r}
client <- doclr_client("https://docling.example.org", api_key = Sys.getenv("MY_KEY"))
doclr_health(client)
```

Both arguments have environment-variable defaults — `DOCLING_SERVE_URL` and
`DOCLING_SERVE_API_KEY` — so in a configured session `doclr_client()` on its
own is usually enough. The key is sent as a redacted `X-Api-Key` header and
never appears in printed requests.

To run a server on your own machine instead, call `doclr_setup()`. It is
explicit and user-invoked on purpose: it asks before installing anything, and
it is never triggered by loading the package.

```{r}
doclr_setup()
```

`doclr_setup()` works through a fallback chain, stopping at the first step
that succeeds:

1. A server already answers on the port — nothing to do.
2. Podman or Docker is on the `PATH` — pull the image and run it.
3. Neither is installed, but Homebrew, winget or apt/dnf/pacman is — offer
   to install Podman, after asking.
4. No container path at all, but Python is present — suggest
   `pip install "docling-serve[ui]"`.
5. None of the above — ask for the URL of a remote server rather than
   guessing. This is the expected path on a locked-down Windows machine with
   no WSL2 and no Python.

Afterwards the container is managed directly:

```{r}
doclr_server_status()
doclr_server_stop()
doclr_server_start()
```

The first pull is large — today's `docling-serve` image bundles model
weights — so expect several hundred megabytes and a slow first run.

## Converting synchronously

The sync endpoints block until the server finishes. They are the right
choice for one small file:

```{r}
doc <- doclr_convert_source("https://arxiv.org/pdf/2408.09869")
cat(as_markdown(doc))
```

Local files are uploaded instead:

```{r}
doc <- doclr_convert_file("reports/quarterly.pdf")
```

## Converting asynchronously

Anything long enough to time out a request belongs on the async endpoints.
`doclr_convert()` wraps the whole cycle — submit, poll with a spinner,
collect — and routes URLs and local paths to the right endpoint for you:

```{r}
doc <- doclr_convert("reports/quarterly.pdf")
```

Drive the steps yourself when you want to do other work in between:

```{r}
task <- doclr_convert_file_async("reports/quarterly.pdf")
task

doclr_task_status(task)

# ... do something else ...

doc <- doclr_wait(task, poll_interval = 5, timeout = 1800)
```

`doclr_task_status()` returns a one-row tibble, which makes it easy to watch
several tasks at once:

```{r}
tasks <- lapply(list.files("reports", full.names = TRUE), doclr_convert_file_async)
do.call(rbind, lapply(tasks, doclr_task_status))
```

A task that fails raises a `doclr_error_task` condition, and one that outlives
its `timeout` raises `doclr_error_timeout` — the job is still on the server,
so you can call `doclr_wait()` again with the same id.

## Choosing what the server produces

`doclr_options()` builds the options object. Only the fields you set are
sent, so server defaults apply to everything else:

```{r}
options <- doclr_options(
  to_formats = c("md", "json"),
  do_ocr = TRUE,
  ocr_lang = c("eng", "fra"),
  do_table_structure = TRUE,
  table_mode = "accurate"
)

doc <- doclr_convert("reports/scanned.pdf", options)
```

Fields this package does not name explicitly pass straight through:

```{r}
doclr_options(to_formats = "md", pdf_backend = "dlparse_v4")
```

## Reading tables and figures out of the result

Markdown is the readable form; JSON is the structured one. Ask for both when
you want text to read and a tree to query:

```{r}
doc <- doclr_convert(
  "reports/quarterly.pdf",
  doclr_options(to_formats = c("md", "json"))
)

cat(as_markdown(doc))

tree <- as_json(doc)
names(tree)
```

The docling JSON tree carries `texts`, `tables` and `pictures` collections.
Tables arrive as grids of cells, which you can flatten into a data frame:

```{r}
tables <- tree$tables

table_to_df <- function(table) {
  grid <- table$data$grid
  rows <- lapply(grid, function(row) vapply(row, function(cell) cell$text, character(1)))
  body <- do.call(rbind, rows[-1])
  df <- as.data.frame(body, stringsAsFactors = FALSE)
  names(df) <- rows[[1]]
  df
}

table_to_df(tables[[1]])
```

Figures live in `tree$pictures`. Ask for embedded images when you want the
bytes rather than a placeholder:

```{r}
doc <- doclr_convert(
  "reports/quarterly.pdf",
  doclr_options(to_formats = "json", image_export_mode = "embedded")
)

pictures <- as_json(doc)$pictures
length(pictures)
```

## The other representations

```{r}
as_text(doc)     # plain text
as_html(doc)     # HTML
as_doctags(doc)  # DocTags markup
```

Each generic reads a representation the server actually produced. Asking for
one you did not request raises a `doclr_error_format` condition naming what
is available — add the format to `to_formats` and convert again.

## Handling failures

Every error is an `rlang` condition with a `doclr_error_*` class, so you can
catch precisely the case you want:

```{r}
tryCatch(
  doclr_convert("reports/quarterly.pdf"),
  doclr_error_connection = function(e) {
    message("No server is running. Try doclr_setup().")
    NULL
  },
  doclr_error_http = function(e) {
    message("The server rejected the request: ", conditionMessage(e))
    NULL
  }
)
```
